When Subgraph Isomorphism is Really Hard, and Why This Matters for Graph Databases

### 2018-02-23 paper/paper.tex
Times

Rather than the modern techniques used by Glasgow and LAD, we deliberately select a very simple
starting point: a constraint programming model implemented in Gecode 5.1.0 using only
toolkit-provided constraints and heuristics. The model uses a ``smallest domain first'' variable
ordering heuristic with tie-breaking on degree, an all-different constraint, extensional constraints
for adjacency, and labelled-degree filtering at the top of search. Even this simple approach finds
nearly every instance in each of the four datasets trivial, and no exponential behaviour is
observed. As we show in \cref{figure:biiiig-data}, the \emph{hardest} instance for the AIDS dataset
requires 287 recursive calls, and 149,268 of the instances can be solved without search.  For PDBS,
the hardest problem requires 3,161 recursive calls (which occurs 20 times, due to duplicated
queries). For both PCM and PPI, the hardest problem requires 23 recursive calls.  In other words,
none of these instances are in any way computationally hard even for a simple domain-based
algorithm, even before introducing stronger filtering, inference, or heuristics.  Furthermore it is
certainly not the case that a sequential scan is infeasible as so many filter~/ verify papers claim.
In contrast, with VF2, the hardest instance in PDBS requires 31,189 recursive calls; in AIDS,
168,569 calls; in PCMS, 6,476,072 calls; and in PPI, 2,913. In other words, any exponential
behaviour seen in these instances is purely down to the choice of subgraph isomorphism algorithm,
and these instances are not inherently ``really hard''.

@@

We do not claim that the ultimate ``big data'' subgraph matching algorithm already exists:
on the contrary, there is likely to be plenty of room for future improvement now that we understand
the importance of getting algorithm design right.  For example, a major disadvantage of using
domains is the relatively expensive initialisation costs, which quickly add up when dealing with
large numbers of trivial instances. Employing a presolver is an obvious approach---and VF2 is
in fact good in this role \citep{DBLP:conf/lion/KotthoffMS16}---but there are other possibilities.
For example, minimal or lazy forward-checking
\citep{DBLP:conf/ictai/DentM94,DBLP:conf/cp/BacchusG95,o:Dent96,DBLP:conf/ecai/LarrosaM98} avoids
constructing every domain upfront, although adopting this require alternatives to ``smallest domain
first'' and to all-different. Such an approach may also be beneficial for huge target graphs, where
having domains range over the entire target is impractical.

>>>>

Rather than the modern techniques used by Glasgow and LAD, we deliberately select a very simple
starting point: a constraint programming model implemented in Gecode 5.1.0 using only
toolkit-provided constraints and heuristics. The model uses a ``smallest domain first'' variable
ordering heuristic with tie-breaking on degree, an all-different constraint, extensional constraints
for adjacency, and labelled-degree filtering at the top of search. Even this simple approach finds
nearly every instance in each of the four datasets trivial, and no exponential behaviour is
observed. As we show in \cref{figure:biiiig-data}, the \emph{hardest} instance for the AIDS dataset
requires 287 recursive calls, and 149,268 of the instances can be solved without search;
furthermore, no instance individually requires more than 12ms. For both PCM and PPI, the hardest
problem requires 23 recursive calls, and no instance takes more than 12ms and 282ms respectively.
For PDBS, the hardest problem requires 3,161 recursive calls (which occurs 20 times, due to
duplicated queries); however, Gecode struggles on these instances due to its general purpose data
structures not scaling well to the large target graph sizes, and some of these instances can take as
long as 30s to solve.

In other words, none of these instances are in any way computationally hard even for a simple
domain-based algorithm, even before introducing stronger filtering, inference, or heuristics, and
the only difficulties arise from Gecode requiring the use of extensional constraints rather than the
specialised adjacency data structures used by Glasgow and LAD.  Furthermore it is certainly not the
case that a sequential scan is infeasible as so many filter~/ verify papers claim.  In contrast,
with VF2, the hardest instance in PDBS requires 31,189 recursive calls; in AIDS, 168,569 calls; in
PCMS, 6,476,072 calls; and in PPI, 2,913. In other words, any exponential behaviour seen in these
instances is purely down to the choice of subgraph isomorphism algorithm, and these instances are
not inherently ``really hard''.

@@

We do not claim that the ultimate ``big data'' subgraph matching algorithm already exists:
on the contrary, there is likely to be plenty of room for future improvement now that we understand
the importance of getting algorithm design right.  For example, a major disadvantage of using
domains is the relatively expensive initialisation costs, which quickly add up when dealing with
large numbers of trivial instances. Employing a presolver is an obvious approach---and VF2 is
in fact good in this role \citep{DBLP:conf/lion/KotthoffMS16}---but there are other possibilities.
For example, minimal or lazy forward-checking
\citep{DBLP:conf/ictai/DentM94,DBLP:conf/cp/BacchusG95,o:Dent96,DBLP:conf/ecai/LarrosaM98} avoids
constructing every domain upfront, although adopting this require alternatives to ``smallest domain
first'' and to all-different. Such an approach may also be beneficial for huge target graphs, where
having domains range over the entire target is impractical. Indeed, we saw that a raw constraint
programming approach using Gecode suffered from relatively long runtimes on the very large but
sparse PDBS instances, despite finding the instances very easy in terms of number of recursive
calls. An algorithm which is capable of performing most of the reasoning of the Glasgow and LAD
solvers, but without needing to store domains and adjacency matrices, could represent a major
breakthrough.
